Manuscript draft

### 2019-09-09 manuscript/manuscript draft.Rmd
updated manuscript draft

---
title: "Global patterns in forest productivity"
author: "Rebecca Banbury Morgan, Valentine Herrmann, Norbert Kunert, Ben Bond-Lamberty, Helene Muller-Landau, [MORE??-CHECK], Kristina J. Anderson-Teixeira"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [library.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl
header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[table]{font=footnotesize}
                  \captionsetup[figure]{font=footnotesize}

@@

Globally, forests play an important role in the carbon cycle and are a major component of global carbon dioxide budgets [@luyssaert_old-growth_2008]. They show higher levels of productivity than non-forest terrestrial ecosystems [@del_grosso_global_2008], and as a result achieve significant carbon sequestration and storage. Estimating the total role of forests in the carbon cycle is challenging, but studies indicate that old growth forests alone sequester up to 1.4 GtCyr^-1^ [@malhi_carbon_1999], while the total sequestration of carbon by established forests globally could be up to 2.4 GtCyr^-1^, with the largest sinks being in old-growth tropical forests [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, with consequences for global climate, there is increasing recognition that proper protection and management of forest resources will have an important role to play in mitigating climate change. Understanding the patterns of forest productivity on a global scale, and the drivers behind them, is therefore a priority in forest research.  [**KAT-work on this; consider shifting top focus from global C cycle to centrality of forest productivity to ecology & climate science.** ] Given the importance of understanding broad-scale patterns in forest productivity, there have been many meta-analyses on the theme. However, our ability to draw general macroscopic conclusions regarding global variation in multiple productivity variables with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables (*e.g.*, combining net primary productivity records with and without belowground components); and typically consider only one or a few variables at a time. The recent development of a global forest carbon database synthesizing multiple variables and including records of stand history (ForC; [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018]) opens up the possibility for a standardized analysis of global scale variation in multiple components of forest productivity and the principle climatic drivers of these patterns.

On a global scale, the productivity of forests varies with latitude, showing a general trend of decreasing productivity with latitude [@beer_terrestrial_2010; @jung_global_2011]. Studies agree that productivity is lowest in the boreal regions, and increases into the temperate regions [@luyssaert_co2_2007; @huston_global_2009; @beer_terrestrial_2010; @jung_global_2011]. However, evidence is inconclusive on whether productivity continues to increase into the tropics, or whether it plateaus in temperate regions. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity (GPP) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011]. This is corroborated by analysis of site-level GPP measurements, which appear to reach their highest levels in tropical forests [@luyssaert_co2_2007]. In contrast, there is evidence that the highest values of net primary productivity (NPP) may be found in temperate forests [@luyssaert_co2_2007; @huston_global_2009], although other studies find NPP is highest in the tropics, showing a decrease with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity (ANPP), finding evidence of a weak negative relationship between ANPP and latitude [@huston_global_2009; @gillman_latitude_2015].
An understanding of the global patterns of productivity is important; however elucidating the drivers of these patterns will be even more valuable in expanding our understanding of global carbon cycling. Primary productivity can be influenced by many factors, which often act across a range of scales, and may show interactive effects with each other. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], management [@simova_enigma_2017]; nutrient availability [@aragao_above-_2009]; and altitude [@girardin_net_2010; @malhi_variation_2017] all impact forest productivity. On a global scale, we expect that productivity is most strongly influenced by broad climatic gradients.

We know from previous research that climate is a significant driver of productivity across broad spatial scales [@cleveland_relationships_2011]. The majority of studies have focused on exploring the relationships between productivity and mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. These variables have the advantage that they describe broad trends in temperature and water availability, and therefore capture a lot of global-scale variation in climate. There is strong evidence that both MAT and MAP show significant positive relationships with productivity [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of productivity across studies. There is support across multiple studies for the hypothesis that various measures of primary productivity saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co2_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity are less conclusive. Luyssaert et al. [-@luyssaert_co2_2007] examined GPP and NPP and found that, while GPP increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in GPP saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that NPP increases linearly with temperature.

The influence of these climate variables on productivity is further complicated by the possibility of interactive effects occurring between them. Taylor et al. [-@taylor_temperature_2017] showed that increased MAP had a negative effect at low MAT, but a positive effect at high levels of MAT, and vice versa. Similarly, Luyssaert et al. [-@luyssaert_co2_2007] showed that ecosystems limited by low MAP do not benefit from increasing MAT, and vice versa, suggesting that both temperature and water availability are important in explaining productivity.

@@

Analyses were conducted on data contained in the open-access ForC database [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018]. This database contains records of field-based measurements of forest carbon stocks and annual fluxes, compiled from original publications and existing data compilations and databases. Associated data, such as stand age, measurement methodologies, and disturbance history, are also included. The database was significantly expanded since the publication of [@anderson-teixeira_forc_2018] through integration with the Global Database of Soil Respiration (SRDB; **Bond-Lamberty, DOI: 10.5194/bg-7-1915-2010**). Additional targeted literature searches were conducted to identify any further available data on primary productivity, with particular focus on mature forests in temperate and boreal regions. ForC currently contains **UPDATE STATS** 29768 records from 5227 plots, representing 20 distinct ecozones across all forested biogeographic and climate zones. We used ForC v.XX, archived on Zenodo with DOI XXX.

>>>>

---
title: "Manuscript draft"
author: "Rebecca Banbury Morgan, Valentine Herrmann, Norbert Kunert, Ben Bond-Lamberty, Helene Muller-Landau, [MORE??-CHECK], Kristina J. Anderson-Teixeira"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
bibliography: [library.bib,  packages.bib]
biblio-style: apalike
csl: apa.csl
header-includes:  \usepackage{float}
                  \usepackage{caption}
                  \captionsetup[table]{font=footnotesize}
                  \captionsetup[figure]{font=footnotesize}

@@

Globally, forests play an important role in the carbon cycle and are a major component of global carbon dioxide budgets [@luyssaert_old-growth_2008]. They show higher levels of productivity than non-forest terrestrial ecosystems [@del_grosso_global_2008], and as a result achieve significant carbon sequestration and storage. Estimating the total role of forests in the carbon cycle is challenging, but studies indicate that old growth forests alone sequester up to 1.4 GtCyr^-1^ [@malhi_carbon_1999], while the total sequestration of carbon by established forests globally could be up to 2.4 GtCyr^-1^, with the largest sinks being in old-growth tropical forests [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, with consequences for global climate, there is increasing recognition that proper protection and management of forest resources will have an important role to play in mitigating climate change. Understanding the patterns of forest productivity on a global scale, and the drivers behind them, is therefore a priority in forest research.  [**KAT-work on this; consider shifting top focus from global C cycle to centrality of forest productivity to ecology & climate science.** ] Given the importance of understanding broad-scale patterns in forest productivity, there have been many meta-analyses on the theme. However, our ability to draw general macroscopic conclusions regarding global variation in multiple productivity variables with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables (*e.g.*, combining net primary productivity records with and without belowground components); and typically consider only one or a few variables at a time. The recent development of a global forest carbon database synthesizing multiple variables and including records of stand history (ForC; [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc:_2018]) opens up the possibility for a standardized analysis of global scale variation in multiple components of forest productivity and the principle climatic drivers of these patterns.

On a global scale, the productivity of forests varies with latitude, showing a general trend of decreasing productivity with latitude [@beer_terrestrial_2010; @jung_global_2011]. Studies agree that productivity is lowest in the boreal regions, and increases into the temperate regions [@luyssaert_co_2007; @huston_global_2009; @beer_terrestrial_2010; @jung_global_2011]. However, evidence is inconclusive on whether productivity continues to increase into the tropics, or whether it plateaus in temperate regions. Evidence for this is further complicated by the fact that different studies use different measures of productivity to explore these relationships. For example, modelling of global terrestrial ecosystem gross primary productivity (GPP) through upscaling and calibration of eddy flux measurements indicates that GPP peaks in tropical forests [@beer_terrestrial_2010; @jung_global_2011]. This is corroborated by analysis of site-level GPP measurements, which appear to reach their highest levels in tropical forests [@luyssaert_co_2007]. In contrast, there is evidence that the highest values of net primary productivity (NPP) may be found in temperate forests [@luyssaert_co_2007; @huston_global_2009], although other studies find NPP is highest in the tropics, showing a decrease with latitude [@simova_enigma_2017]. Other studies have chosen to focus exclusively on above-ground net primary productivity (ANPP), finding evidence of a weak negative relationship between ANPP and latitude [@huston_global_2009; @gillman_latitude_2015].
An understanding of the global patterns of productivity is important; however elucidating the drivers of these patterns will be even more valuable in expanding our understanding of global carbon cycling. Primary productivity can be influenced by many factors, which often act across a range of scales, and may show interactive effects with each other. On a local scale, stand age [@litton_carbon_2007; @gillman_latitude_2015], management [@simova_enigma_2017]; nutrient availability [@aragao_above-_2009]; and altitude [@girardin_net_2010; @malhi_variation_2017] all impact forest productivity. On a global scale, we expect that productivity is most strongly influenced by broad climatic gradients.

We know from previous research that climate is a significant driver of productivity across broad spatial scales [@cleveland_relationships_2011]. The majority of studies have focused on exploring the relationships between productivity and mean annual temperature (MAT) and mean annual precipitation (MAP), as the most commonly reported site-level climate variables. These variables have the advantage that they describe broad trends in temperature and water availability, and therefore capture a lot of global-scale variation in climate. There is strong evidence that both MAT and MAP show significant positive relationships with productivity [@chu_does_2016]. However, as with latitude, the shape of those relationships is not always clear, and, again, is complicated by the use of different measures of productivity across studies. There is support across multiple studies for the hypothesis that various measures of primary productivity saturate at high levels of MAP, though the saturation points identified vary from 1500mm [@luyssaert_co_2007] up to 2445mm MAP [@schuur_productivity_2003]. Studies of the influence of MAT on productivity are less conclusive. Luyssaert et al. [-@luyssaert_co_2007] examined GPP and NPP and found that, while GPP increases linearly with MAT, NPP saturates at around 10$^\circ$C MAT. In contrast, Larjavaara and Muller-Landau [-@larjavaara_temperature_2012], find that increases in GPP saturate at approximately 25$^\circ$C MAT, while Schuur [-@schuur_productivity_2003] shows that NPP increases linearly with temperature.

The influence of these climate variables on productivity is further complicated by the possibility of interactive effects occurring between them. Taylor et al. [-@taylor_temperature_2017] showed that increased MAP had a negative effect at low MAT, but a positive effect at high levels of MAT, and vice versa. Similarly, Luyssaert et al. [-@luyssaert_co_2007] showed that ecosystems limited by low MAP do not benefit from increasing MAT, and vice versa, suggesting that both temperature and water availability are important in explaining productivity.

@@

Analyses were conducted on data contained in the open-access ForC database [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc:_2018]. This database contains records of field-based measurements of forest carbon stocks and annual fluxes, compiled from original publications and existing data compilations and databases. Associated data, such as stand age, measurement methodologies, and disturbance history, are also included. The database was significantly expanded since the publication of [@anderson-teixeira_forc:_2018] through integration with the Global Database of Soil Respiration (SRDB; **Bond-Lamberty, DOI: 10.5194/bg-7-1915-2010**). Additional targeted literature searches were conducted to identify any further available data on primary productivity, with particular focus on mature forests in temperate and boreal regions. ForC currently contains **UPDATE STATS** 29768 records from 5227 plots, representing 20 distinct ecozones across all forested biogeographic and climate zones. We used ForC v.XX, archived on Zenodo with DOI XXX.
